---
title: "M3: Class Exercise on Trend and Season"
author: "Luana Lima"
output: pdf_document
---

## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)

```

## Importing data

Let's continue working with our inflow data for reservoirs in Brazil. 

```{r}

#Importing time series data from text file#
raw_inflow_data <- read.table(file="../Data/inflowtimeseries.txt",header=FALSE,skip=0)

#Trim the table to include only columns you need
nhydro <- ncol(raw_inflow_data)-2
nobs <- nrow(raw_inflow_data) 

#If your file does not have header like this one you can add column names after 
#creating the data frame
colnames(raw_inflow_data)=c("Month","Year", "HP1", "HP2","HP3","HP4", "HP5",
                            "HP6","HP7", "HP8","HP9","HP10", "HP11","HP12", 
                            "HP13", "HP14","HP15")

#Checking data
head(raw_inflow_data)
str(raw_inflow_data)
```
## Creating the date object

Here we use the function my() from package lubridate.

```{r}
#using package lubridate
my_date <- paste(raw_inflow_data[,1],raw_inflow_data[,2],sep="-")
my_date <- my(my_date)  #function my from package lubridate
head(my_date)
#add that to inflow_data and store in a new data frame 
inflow_data <- cbind(my_date,raw_inflow_data[,3:(3+nhydro-1)])
head(inflow_data)

```


## Initial Plots

Initial time series plot.

```{r}
#using package ggplot2
for(i in 1:nhydro){
  print(ggplot(inflow_data, aes(x=my_date, y=inflow_data[,(1+i)])) +
            geom_line(color="blue") +
            ylab(paste0("Inflow ",colnames(inflow_data)[(1+i)],sep="")) 
        )
}

```

## Zeros in the end on data 

The initial plots showed that we have zeros in the end of the data set. It could be missing observation or observation that haven't been observed yet. Use the tail() to find out how many zeros you have and how many lines you will need to remove.

```{r}
#check the final obs on data 
tail(inflow_data)


```

Note our last observation is from August 2011 but the data file was filled with zeros. Let's remove the last four rows of our data set.

```{r}
#Remove last year by replacing current data frame
inflow_data <- inflow_data[1:(nobs-12),]

#update object with number of observations
nobs <- nobs-12

#Tail again to check if the rows were correctly removed
tail(inflow_data)
```
Fixed!

## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_inflow_data <- ts(inflow_data[,2:(2+nhydro-1)],start=c(1931,1),frequency=12) 
#note that we are only transforming columns with inflow data, not the date columns  #start=my_date[1],end=my_date[nobs],
head(ts_inflow_data,12)
```


## Plotting ACF and PACF

Let's use functions Acf() and Pacf() from package "forecast".

```{r acf}
#Acf and Pacf for HP1
for(i in 1:nhydro){
  par(mfrow=c(1,2))  #place plot side by side
  Acf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep="")) 
  # because I am not storing Acf() into any object, I don't need to specify plot=TRUE 
  Pacf(ts_inflow_data[,i],lag.max=40,main=paste("Inflows HP",i,sep=""))
}

```


## Trend Component - Linear Model

Let's identify and remove trend component. You start by fitting a linear model to $Y_t=\beta_0+\beta_1*t+\epsilon_t$. 

```{r}
#Create vector t
t <- 1:nobs

#for(iHP in 1:nhydro){}
iHP <- 4 #HP4

#Fit a linear trend to TS of iHP
linear_trend <- lm(inflow_data[,(iHP+1)]~t)
summary(linear_trend)

```


```{r}
#remove the trend from series
beta0 <- linear_trend$coefficients[1]
beta1 <- linear_trend$coefficients[2]

y_detrend <- inflow_data[,iHP+1] - (beta0 + beta1*t)

df_detrend <- data.frame("date"=inflow_data$my_date,
                         "observed"=inflow_data[,iHP+1],
                         "detrend"=y_detrend)
ggplot(df_detrend,aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")

```

Note that blue line is our original series, red line is our trend, green line is our original series minus the trend or in other words the detrended series. And in orange is the trend line for the detrended series which has slope 0 meaning we were able to effectively eliminate the trend with a linear model.

## Check what happened to ACF after removing the linear trend

```{r}
library(cowplot) #for plot_grid
library(ggfortify) #for autoplot()

ts_y_detrend <- ts(y_detrend,start=c(1931,1),frequency=12)

plot_grid(
  autoplot(ts_inflow_data[,iHP],ylab="Observed Inflow"),
  autoplot(Acf(ts_inflow_data[,iHP],lag.max=40,plot=FALSE),main=NULL),
  autoplot(Pacf(ts_inflow_data[,iHP],lag.max=40,plot=FALSE),main=NULL),
  autoplot(ts_y_detrend,ylab="Detrended Inflow - Linear Model"),
  autoplot(Acf(ts_y_detrend,lag.max=40,plot=FALSE),main=NULL),
  autoplot(Pacf(ts_y_detrend,lag.max=40,plot=FALSE),main=NULL),
  nrow=2,ncol=3
)
```

## Check what happened to ACF after removing the trend using moving average




## Seasonal Component

Now let's shift attention to the seasonal component.

```{r}
#Use seasonal means model
iHP <- 4

#create dummies
dummies <- seasonaldummy(ts_inflow_data[,iHP])

#regress on dummies
seas_linear_model <- lm(inflow_data[,(iHP+1)]~dummies)
summary(seas_linear_model)

#store coeff
beta0 <- seas_linear_model$coefficients[1]
beta1 <- seas_linear_model$coefficients[2:12]

#seasonal comp
seas_comp <- array(0,nobs)
for(i in 1:nobs){
  seas_comp[i] <- beta0 + beta1 %*% dummies[i,]
}

ggplot(inflow_data,aes(x=my_date))+
  geom_line(aes(y=inflow_data[,(iHP+1)]),col="black")+
  geom_line(aes(y=seas_comp),col="orange")

y_deseason <- inflow_data[,(iHP+1)] - seas_comp

ggplot(inflow_data,aes(x=my_date))+
  geom_line(aes(y=inflow_data[,(iHP+1)]),col="black")+
  geom_line(aes(y=y_deseason),col="green")


```

## Check what happened to ACF after removing the linear trend

```{r}
library(cowplot) #for plot_grid
library(ggfortify) #for autoplot()

ts_y_deseason <- ts(y_deseason,start=c(1931,1),frequency=12)

plot_grid(
  autoplot(ts_inflow_data[,iHP],ylab="Observed Inflow"),
  autoplot(Acf(ts_inflow_data[,iHP],lag.max=40,plot=FALSE),main=NULL),
  autoplot(Pacf(ts_inflow_data[,iHP],lag.max=40,plot=FALSE),main=NULL),
  autoplot(ts_y_deseason,ylab="Deseasonalized Inflow"),
  autoplot(Acf(ts_y_deseason,lag.max=40,plot=FALSE),main=NULL),
  autoplot(Pacf(ts_y_deseason,lag.max=40,plot=FALSE),main=NULL),
  nrow=3,ncol=3
)
```
## decompose

```{r}
iHP <- 1
plot(decompose(ts_inflow_data[,iHP]))
```


##Exercise

Fit trend and seasonal for the other variables HP2, HP3, ...


## Stationarity Tests in R

Some test only work for non-seasonal data. So let's create another series with yearly averages for inflow.

```{r}
#Group data in yearly steps instances


```

### Mann Kendall 

Check for deterministic trend.

```{r}

#Since I have seasonal data I cannot use the simple MannKendall()
#another example of functions that need a ts object


#Use yearly date to run Mann Kendall

```

### Spearman Correlation test 

Also check for deterministic trend, for non-seasonal data only.

```{r}
#Deterministic trend with Spearman Correlation Test
print("Results from Spearman Correlation")


#with cor.test you can get test statistics

```

### Augmented Dickey Fuller (ADF)

Used to detect stochastic trend.

```{r}
#Null hypothesis is that data has a unit root
print("Results for ADF test/n")

#Now let's try the yearly data
print("Results for ADF test on yearly data/n")

```

##Exercise

Run the stationarity tests for the other variables HP2, HP3, ...

